---
title: "Data processing"
author: "Yan Wang"
date: "10/01/2022"
output: 
  github_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(ggplot2)
library(reshape2)
SQR<-read_excel("~/Desktop/DS4Ling-2021/old repository/private/DS4Lingdataset/SQR.xlsx")
OEQ<-read_excel("~/Desktop/DS4Ling-2021/old repository/private/DS4Lingdataset/OEQ- text.xlsx")
Metadata<-read_excel("~/Desktop/DS4Ling-2021/old repository/private/DS4Lingdataset/Metadata of participants.xlsx")
```

# Data set description

Data set 1 SQR: participant ID, time of controllability assessment, symptom sequence, and 3 mean controllability scores

Data set 2 OEQ: participant ID, symptom  and answers for the 5 questions: 

1. What does the symptom make you feel like? 

2. What do you think is causing your symptom? 

3. When did you first notice it, and does it follow any sort of pattern? 

4. How does the symptom affect you?

5. Have you tried anything? Is it helpful?

Data set 3 Metadata: participant ID and sociodemographics, including age, marrige status, employment, education, race, ethnicity. 

# Data preparation

```{r}
names(SQR)
SQR_8BL<- SQR %>% filter(`Administration Number`=="Baseline (week 0)"|`Administration Number`== "8 week f/u")
names(SQR_8BL)[3:9]<-c("Time", "S1Cont", "S2Cont", "S3Cont" , "S1", "S2", "S3")
names(SQR_8BL)
```

## Reshape Df1 because we want one patient has multiple entries for symptoms, baseline contrallability scores, and controllability scores

```{r}
library(tidyverse)
SQR_8BL1<-
  SQR_8BL %>%
  gather("S1","S2","S3",key="SymptomSequence",value="SymptomName") %>% select(-c(4:6))
SQR_8BL2<-
  SQR_8BL %>%
  gather("S1Cont","S2Cont","S3Cont",key="SymptomSequence",value="Controllability") %>% select(-c(4:6))
SQR_8BL2$SymptomSequence<-as.factor(SQR_8BL2$SymptomSequence)
SQR_8BL1$SymptomSequence<-as.factor(SQR_8BL1$SymptomSequence)
levels(SQR_8BL2$SymptomSequence)[1]<-"S1"
levels(SQR_8BL2$SymptomSequence)[2]<-"S2"
levels(SQR_8BL2$SymptomSequence)[3]<-"S3"
levels(SQR_8BL1$SymptomSequence)
head(SQR_8BL1)[3:5]
head(SQR_8BL2)[3:5]
```

```{r}
Df<- inner_join(SQR_8BL1, SQR_8BL2, by= c("Participant ID", "GOGID", "Time", "SymptomSequence")) %>% na.omit()
names(Df)[1]<-"ID"
```

## Merge with OEQ (text data) by participant ID and Symptom
```{r}
names(OEQ)[1:2]<-c("ID", "SymptomName")
Df8BL<-left_join(Df ,OEQ, by=c("ID", "SymptomName")) %>% na.omit()
names(Df8BL)[7:9]<-c("FeelingCause", "Effect", "Strategy")
```

## Save the output for the linguisic feature extraction
```{r}
write.csv(Df8BL, "~/Desktop/Df8BL for text analysis.csv")
```

