---
title: "SD controllability Analysis"
author: "Yan Wang"
date: "10/24/2021"
output:
  github_document:
    toc: yes
    
---

# Load all datasets
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
library(readxl)
library(dplyr)
library(ggplot2)
library(reshape2)
SQR<-read_excel("~/Desktop/DS4Ling-2021/old repository/private/DS4Lingdataset/SQR.xlsx")
OEQ<-read_excel("~/Desktop/DS4Ling-2021/old repository/private/DS4Lingdataset/OEQ- text.xlsx")
Metadata<-read_excel("~/Desktop/DS4Ling-2021/old repository/private/DS4Lingdataset/Metadata of participants.xlsx")
```
# Data preparation

## Calculate the dependent variable = the difference of symptom controllability scores at baseline and 8 weeks (at the end of the intervention)

```{r}
SQR_8<- SQR %>% filter(`Administration Number`== "8 week f/u")
SQR_BL<- SQR %>% filter(`Administration Number`=="Baseline (week 0)")
names(SQR_8)[3:9]<-c("Time", "S1Cont", "S2Cont", "S3Cont" , "S1", "S2", "S3")
names(SQR_BL)[3:9]<-c("Time", "S1Cont", "S2Cont", "S3Cont" , "S1", "S2", "S3")
names(SQR_BL)
names(SQR_8)
sum(is.na(SQR_BL$S1Cont))
sum(is.na(SQR_8$S1Cont))
SQR_8BL<-inner_join(SQR_8, SQR_BL, by=c("Participant ID", "GOGID", "S1", "S2", "S3"))
names(SQR_8BL)
SQR_8BL$S1change<-SQR_8BL$S1Cont.x-SQR_8BL$S1Cont.y
SQR_8BL$S2change<-SQR_8BL$S2Cont.x-SQR_8BL$S2Cont.y
SQR_8BL$S3change<-SQR_8BL$S3Cont.x-SQR_8BL$S3Cont.y
names(SQR_8BL)
df1<-SQR_8BL %>% dplyr::select(1,2, 7:9, 14:16)
names(df1)
names(df1)[1:8]<-c("ID", "GOGID", "S1", "S2", "S3", "8BLS1change","8BLS2change", "8BLS3change" )
names(df1)
```

## Reshape df1 because we want one patient has multiple entries for symptoms and controllability scores

```{r}
df1_1<-df1 %>% melt(id.vars = c("ID", "GOGID"), measure.vars =c("S1", "S2", "S3"), variable.name = "SymptomNo", value.name = "Symptom") 
names(df1_1)
df1_2<-df1 %>% melt(id.vars = c("ID", "GOGID"), measure.vars =c( "8BLS1change", "8BLS2change", "8BLS3change"), variable.name = "toy", value.name = "8BLChange")
colnames(df1_1)
colnames(df1_2)
levels(df1_2$toy)[1]<-"S1"
levels(df1_2$toy)[2]<-"S2"
levels(df1_2$toy)[3]<-"S3"
names(df1_2)[3]<-"SymptomNo"
names(df1_2)
df1_3<- inner_join(df1_1, df1_2, by= c("ID", "GOGID", "SymptomNo"))
df1_4<-df1_3 %>% na.omit()
names(df1_4)
```

## Merge with text data by Participant ID and Symptom
```{r}
#install.packages("splus2R")
library(splus2R)
names(OEQ)[1:2]<-c("ID", "Symptom")
df8BL<-left_join(df1_4,OEQ, by=c("ID", "Symptom")) %>% na.omit()
names(df8BL)
names(df8BL)[6:8]<-c("feelingCause", "impact", "strategies")
```

# Descriptive stats of the controllability score changes from baseline to 8 wks
## Distribution - normality assumed
```{r}
ggplot(df8BL)+ 
  geom_histogram(binwidth=0.05, color="blue",aes(x=`8BLChange`, y=..density.., fill=..count..))+ 
  stat_function(fun=dnorm,color="blue",
                args=list(mean=mean(df8BL$`8BLChange`),sd=sd(df8BL$`8BLChange`)))
#QQplots
qq<-data.frame(c(df8BL,qqnorm(df8BL$`8BLChange`)))
ggplot(qq,aes(x=x,y=y,legend.position="none"))+
  geom_point()+
  geom_smooth(method="lm")+
  labs(title="Q-Q Normal Plot",x="Theoretic",y="Observed")+
  theme_bw()
##boxplots
ggplot(df8BL)+
  geom_boxplot(aes(df8BL$`8BLChange`))+
  theme_bw()+
  theme(legend.position="none")
shapiro.test(df8BL$`8BLChange`)
#install.packages("pastecs")
library(pastecs)
stat.desc(df8BL$`8BLChange`)
quantile(df8BL$`8BLChange`, 0.75)-quantile(df8BL$`8BLChange`, 0.25)
table(df8BL$`8BLChange`)
```
## Dichotomized controllability score changes into improved vs not improved
```{r}
df8BL$`8BLRecode`[df8BL$`8BLChange`<=0]<-"Not improved"
df8BL$`8BLRecode`[df8BL$`8BLChange`>0]<-"Improved"
```
